---
title: "What `aes()` Really Does in ggplot2"
author: "Ajit Khanal"
format:
  html:
    toc: false
comments:
  giscus:
    repo: AjitKhanal8/ajitkhanal8.github.io
    repo-id: R_kgDOQCigOw
    category: Announcements
    category-id: DIC_kwDOQCigO84DFA7j
    mapping: pathname
    reactions-enabled: true
    input-position: top
    theme: light
    language: en
    loading: lazy
execute:
  echo: true
  warning: false
  message: false
---

In today's blog, I want to briefly explain what `aes()` does in `ggplot()`. There are two main takeaways:

- `aes()` is a function that maps variables in your dataset, or variables you create after transformation, to a **new dataset: a visual dataset**, which `geoms` then use to construct the plot.
- All the variables that you feed into `aes()` must have the same number of rows.

Understanding these two things helps build a mental model before making a plot, and it also means you do not have to memorize all the arguments that `aes()` can take.

Everything before this point is the big idea. Now let us work through it with a real dataset that comes with `ggplot2`.

Load the necessary library. We will use `ggplot2` for plotting.

```{r}
library(ggplot2)
```

## Start with one built-in dataset

We will use `mpg`, a dataset that comes with `ggplot2`.

This dataset records different cars and their fuel economy. In the examples below, we will mainly use:

- `displ`: engine displacement
- `hwy`: highway miles per gallon
- `class`: the type of car
- `drv`: whether the car is front-wheel, rear-wheel, or four-wheel drive

| manufacturer | displ | hwy | class   | drv |
|--------------|------:|----:|---------|-----|
| audi         |   1.8 |  29 | compact | f   |
| audi         |   1.8 |  29 | compact | f   |
| audi         |   2.0 |  31 | compact | f   |
| audi         |   2.0 |  30 | compact | 4   |
| audi         |   2.8 |  26 | compact | 4   |
| audi         |   2.8 |  26 | compact | 4   |

This is only a preview of the first 6 rows. The full dataset has `r nrow(mpg)` rows.

## The simple idea

`aes()` takes columns from your dataset and gives them new jobs in the plot. It is useful to think of this as making a small visual dataset that the geom can use later.

## Example 1: map two columns to `x` and `y`

Suppose we write:

```{r, eval=FALSE}
ggplot(mpg, aes(x = displ, y = hwy))
```

Before we even add a geom, `aes()` is already doing something useful. It is matching:

- `displ` to `x`
- `hwy` to `y`

So the visual dataset begins like this:

Please do not worry about how to make this table. For this blog, just focus on understanding what `aes()` does.

<table class="table table-sm table-striped">
  <thead>
    <tr>
      <th>x</th>
      <th>y</th>
    </tr>
  </thead>
  <tbody>
    <tr><td>1.8</td><td>29</td></tr>
    <tr><td>1.8</td><td>29</td></tr>
    <tr><td>2.0</td><td>31</td></tr>
    <tr><td>2.0</td><td>30</td></tr>
    <tr><td>2.8</td><td>26</td></tr>
    <tr><td>2.8</td><td>26</td></tr>
  </tbody>
</table>

Nothing has been drawn yet. We just have a new table that says where each row should go on the plot.

Then when we add a geom:

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point()
```

## Example 2: add a transformation

Now suppose we transform `displ` inside `aes()`:

```{r, eval=FALSE}
ggplot(mpg, aes(x = log(displ), y = hwy))
```

This still works because `log(displ)` returns one value for each row of the dataset.

So the visual dataset still has the same number of rows. It just begins like this:

<table class="table table-sm table-striped">
  <thead>
    <tr>
      <th>x</th>
      <th>y</th>
    </tr>
  </thead>
  <tbody>
    <tr><td>0.588</td><td>29</td></tr>
    <tr><td>0.588</td><td>29</td></tr>
    <tr><td>0.693</td><td>31</td></tr>
    <tr><td>0.693</td><td>30</td></tr>
    <tr><td>1.030</td><td>26</td></tr>
    <tr><td>1.030</td><td>26</td></tr>
  </tbody>
</table>

Then we can draw the plot:

```{r}
ggplot(mpg, aes(x = log(displ), y = hwy)) +
  geom_point()
```

This example helps explain the second takeaway: transformed variables are fine inside `aes()` as long as they still give one value per row.

## Example 3: create a new visual variable inside `aes()`

In this example, the mapping uses a logical statement.

Suppose we write:

```{r, eval=FALSE}
ggplot(mpg, aes(x = displ, y = hwy, color = hwy > 30))
```

Now `aes()` is matching:

- `displ` to `x`
- `hwy` to `y`
- `hwy > 30` to `color`

So the visual dataset begins like this:

<table class="table table-sm table-striped">
  <thead>
    <tr>
      <th>x</th>
      <th>y</th>
      <th>color</th>
    </tr>
  </thead>
  <tbody>
    <tr><td>1.8</td><td>29</td><td>FALSE</td></tr>
    <tr><td>1.8</td><td>29</td><td>FALSE</td></tr>
    <tr><td>2.0</td><td>31</td><td>TRUE</td></tr>
    <tr><td>2.0</td><td>30</td><td>FALSE</td></tr>
    <tr><td>2.8</td><td>26</td><td>FALSE</td></tr>
    <tr><td>2.8</td><td>26</td><td>FALSE</td></tr>
  </tbody>
</table>

The new `color` column is made of `TRUE` and `FALSE`, but it still has one value per row.

Then we can draw the plot:

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = hwy > 30)) +
  geom_point()
```

## Example 4: Bonus example

Here is one more example using another argument inside `aes()`:

```{r, eval=FALSE}
ggplot(mpg, aes(x = displ, y = hwy, shape = drv))
```

Here `drv` is being mapped to `shape`, just like `region` could be mapped to shape in a different dataset.

So the visual dataset begins like this:

<table class="table table-sm table-striped">
  <thead>
    <tr>
      <th>x</th>
      <th>y</th>
      <th>shape</th>
    </tr>
  </thead>
  <tbody>
    <tr><td>1.8</td><td>29</td><td>f</td></tr>
    <tr><td>1.8</td><td>29</td><td>f</td></tr>
    <tr><td>2.0</td><td>31</td><td>f</td></tr>
    <tr><td>2.0</td><td>30</td><td>4</td></tr>
    <tr><td>2.8</td><td>26</td><td>4</td></tr>
    <tr><td>2.8</td><td>26</td><td>4</td></tr>
  </tbody>
</table>

And then the geom uses that mapping:

```{r}
ggplot(mpg, aes(x = displ, y = hwy, shape = drv)) +
  geom_point()
```

There are many other arguments that can go inside `aes()`, but the basic idea is the same. At this point, all you need to understand are the two takeaways, which I will repeat here:

- `aes()` is a function that maps variables in your dataset, or variables you create after transformation, to a **new dataset: a visual dataset**, which `geoms` then use to construct the plot.
- All the variables that you feed into `aes()` must have the same number of rows.

## Reference

Wickham, Hadley. 2016. *ggplot2: Elegant Graphics for Data Analysis*. 2nd ed. Use R! Cham: Springer. https://doi.org/10.1007/978-3-319-24277-4

## Note

Codex was used to help refine the examples, shape the visual-dataset explanations, and format the post for the website.
